Quantifying the manageability of pollination networks in an invasion context

### 2017-05-04 paper/response_to_reviewers.Rmd

Furthermore, we see the fact that we focus on $\mathbf{B}$ and the network structure as an advantage rather than a disadvantage. 
As one of the previous reviewers highlighted, our manuscript "...presents to ecologists a novel way of studying the relationship between topology and dynamics that is less dependent (when compared to other methods) on particular choices about how dynamics are modelled". 
It can, therefore, be used to gain insight from different ecological systems and is readily applicable to currently available data.

As it has been acknowledged in the control theory literature, "while many complex systems are characterised by nonlinear interactions between the components, the first step in any control challenge is to establish the controllability of the locally linearized system... Before we can explore the fully nonlinear dynamical setting, which is mathematically much harder, we must understand the impact of the topological characteristics on linear controllability, serving as a prerequisite of nonlinear controllability" [@Liu2016].

Along similar lines to the above, we have rewritten the a couple of paragraphs at the beginning of the "Theoretical framework" subsection (Lines 84-99) where we clearly state the assumption of linearity and how it does not invalidate the gained insight. the following modifications to the manuscript. 
In addition, we hade some minor modifications to the last paragraph of the discussion to make more clear our contribution and how it relates to posible future research (Lines 356-371).

### 2017-05-04 paper/response_to_reviewers.Rmd

    *Thanks for this comment and those from other reviewers which have helped us identify some shortcomings on the way we were introducing the control framework in the paper. In reposnse, we have largely rewritten the 'Theoretical framework' section. We start by offering a brief overview of control theory. In this overview, we emphasise our assumptions regarding the dynamical nature of the interactions. We then follow with a brief discussion that indicates how structural controllability is unable to provide information about the particulars of the control signal and the feasibility of its application. Without disregarding the usefulness of the proposed approach, we also mention at the end of the discussion how recently developed numerical methods [e.g. @Cornelius2013] might offer future avenues results in this regard (Lines 363--369)*

- Furthermore, how do the authors interpret the methods by which cycles are controlled in linear control theory (see e.g. doi 10.1126/science.1242063 and its supplementary material) in the context of ecological management? How frequent are cycles in the authors' ecological networks? I should emphasize that I am a fan of control theory and I think it is interesting to apply it as the authors do in this ms. However, the caveats that come along with the points I outline above need to be made clear for the reader. The authors might be interested in doi 10.1063/1.4931570 for a discussion of some relevant ideas.

    *Thanks for pointing this out. In fact, we explored the implementation of the approach of [@Ruths2014] at initial stages of the project. However, we earlier decided to exclude this approach from the final draft, mainly due to space limitations. Nevertheless, due to this comment and one of Referee 2 we have rediscovered its value and use it to support our decision to include the analysis of superior nodes, and not just driver nodes, in our manuscript. Though space constraints remain and we can only mention it briefly in the main text (Lines 329--331), we also dedicate a whole section in the Supporting Information (S2) to this subject. To summarise here, we found our networks weighted by asymmetry to be external dilation dominated, and hence have only a small number of cycles. Networks weighted by mutual dependences are dominated by cycles and hence challenging to address with our approach. In another section in the Supporting Information (S6), we discuss the implications of using these networks and our approach to circumvent them. Finally we are grateful for the reference to @Motter2015. We were unaware of this paper and it has been extremely useful to contextualise our work in the broader control-theory literature.*

@@

- The approach described in lines 145-152 would be clearer if the authors included a figure demonstrating each step for a simple network.

    *Thanks for pointing this out. Exactly as suggested, we have included a figure that helps explain our method to compute all possible maximum cardinality matchings. Nevertheless, because of space limitations and because the methodology itself is not one of our main contributions, we have included the procedure and the figure in the Supporting Information (S1 and Figure S2).*

- Lines 173-176: "To do so, we calculated the maximum matching of the corresponding pollination network, and estimated the minimum proportion of species that need directed interventions to fully control the species abundances in the community." Why is this an estimate? The number of species that need to be directly controlled can be determined unambiguously.

    *We again apologise for the poor choice of words. As the referee points out it, this is not an estimation but an exact calculation. The sentence (Lines 179--181) now reads "To do so, we calculated the networks' maximum matching and determined the minimum proportion of species *n~D~* that need external input signals to fully control the species abundances in the community".*

@@

- Discussion. McDonaldMadden et al (2016) actually show that network metrics are not useful to prioritise the management of foodweb, i.e. to determine which species should be managed first (if required). Please edit current text and analyse differences with your work.

    *Thank you for this suggestion. We acknowledge that we did a poor job embodying our work in the previous literature. We now include a paragraph in the discussion to explore the similarities and differences of our work with previous indices based on structural metrics. We explain (Lines 350--352) how "our approach... is not based on a single structural metric but instead acknowledges the existence of multiple management strategies. By allowing for the fact that some strategies are better than others depending on the context, control theory implicitly highlights that management decisions should not be based on a single technique". We also concur that more flexible approaches like ours and the one implemented by @McDonald-Madden2016 "that take a network-wide approach might prove more useful to guide ecosystem management if we want to go beyond using network metrics to minimise topological species loss".*

- Discussion. Please provide some insight into how your method/analysis/results would differ for non- pollinator invaded networks.

    *We now brifly mention in the Discussion (Lines 359--361) how "although our pollination specific results might not be directly translatable to other ecological systems because of the degree constrains imposed by bipartite networks, the approach we propose is applicable wherever species abundances are influenced by their interactions...".*

@@

- Discussion. Please discuss the caveats of your approach. I suggest moving some of the statistical analyses in sup info if word limit is an issue.

    *We agree that we did not previously discuss the assumptions and caveats of our approach with enough detail. As such, as we now try to better highlight the rationale behind our methods while being more explicit about the assumptions and limitations. Moreover we now provide a fairer assessment of the extent to which our approach might be useful for practical applications in the discussion. We note, however that the extended discussion (alongside with a better development of the theoretical framework), meant that we had to move some of the details about the statistical analysis to the Supporting Information (as suggested).*

- The maximum matching algorithm is computationally expensive as noted by the authors. In Ruths & Ruths (2014) the authors come with an interesting alternative to calculate the minimum number of independent controls (driver species) using only the number of source nodes, external dilation points and internal dilation points; and a heuristic using source and sink nodes. What is the profile of the pollinator networks studied? Would the Ruths & Ruths approach be an easier way of reaching the same conclusions? Would it bring complementary information?

    *Calculating __one__ maximum matching for ecological networks with a few hundred species is not a computationally expensive operation, and therefore we use directly the algorithm and not the approximation proposed by @Ruths2014. Our most expensive computation was to find __all__ the possible maximum cardinality matchings to calculate species-level metrics. Unfortunately, a network's control profile simply does not provide species-level information and hence we couldn't use it to our advantage. Nevertheless, as suggested here and by another reviewer, we have included a section in the Supporting Information (S2) where we explain that our networks are external dilation dominated. Consequently, the approach of @Ruths2014 has provided us with very useful insight on the selection of the direction of control and the utility of the concept of driver nodes for management applications (Lines 209--212 and 329--331).*

@@

## Referee 3 (ELE-00738-2016)

In this work Cagua et al apply tools rooted in control theory to ecological networks and discuss how this approach could be used for management of natural systems, with emphasis on the effects of invasive species. This is an exciting work. It presents to ecologists a novel way of studying the relationship between topology and dynamics that is less dependent (when compared to other methods) on particular choices about how dynamics are modeled. The authors illustrate the potential of this approach by exploring how invasive species change system dynamics and manageability. My only concern about this manuscript is that the authors need to be more careful and more explicit about some limitations of this approach and the generality of their results, specially considering the implications of their results for management.

- Only one type of interaction is being considered here when species are involved in multiple types of interactions. An assumption that is not explicit in the text is that all the factors controlling population densities are the interactions between plants and pollinators. The boundaries of a network, including the types of interactions that are considered or not, are often arbitrarily defined. However, this approach is highly dependent on the topology of the network and thus on how we define the network. I know this limitation applies to any study on system dynamics, but it becomes more serious when we argue that management decisions can be based on a single technique. I have two suggestions. First, the authors should discuss in detail these assumptions and their limitations.

    *We have now included comments about the general limitations of the structural-controllability approach in the "Theoretical framework" section. Furthermore, we are now explicit about the fact that species have multiple types of interactions and therefore our approach can only provide a partial picture when only one type of interaction is included (Lines 363--364). Finally, we are also explicit about the fact that we do not advocate for management decisions to be based solely on our approach (Lines 352--354).*

@@

- Ln 99. Indicate how the matching is represented in the figure. e.g.,  "(dark grey links in Figure 1a)". Figure 1.Indicate the driver nodes in these networks

    *We have redesigned Figure 1 and now we indicate more clearly how the matching works and specify which are the superior, matched, and unmatched nodes. Because we also illustrate the methods to obtain the relative importance, it is now easier to highlight which are the key species in a simple network.*

- Ln 128. Be clear about the meaning of resources here. I understand you mean financial resources not  resources in the ecological sense.

    *We apologise for the imprecision as we meant financial resources. We have modified the sentence to improve clarity (Lines 129--132) and it now reads "Ecologically, potential differences between species are relevant because management and conservation resources are limited, and therefore ecological interventions should be focused on the set of species that might provide the largest impact".*

@@

- Ln 293: Typo: reference

    *Thanks. We corrected it.*

- Ln 303-305. Please explain here which is the response variable that the ratio of plant to pollinator can explain.

    *We have attempted to eliminate ambiguity by rephrasing the sentence (Lines 267--269). It now reads "... of the various covariates we explored, the ratio of plants to pollinators showed the strongest relationship with n~D~".*

### 2017-06-16 paper/manuscript.Rmd
Changes to the theoretical framework part

Disregarding practical considerations, any network (community) could hypothetically be fully controlled if we control the state of every single node (species) individually.
At the core of control theory of complex networks, however, sits the idea that the state of any given node depends on the state of the nodes it interacts with, and the form of this dependence is determined by both the dynamic relationship among interacting nodes and the structure of the links in the network. 
This principle can be leveraged to find a subset of *driver nodes* to which to apply external input signals which then drive the state of every other node in the network to a desired configuration [@Liu2016].

Conveniently the information necessary to determine the minimum number of driver nodes *D* in a system with linear dynamics is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how the species respond to the external input, respectively. 
If *S* is the number of species in the community, the matrix **A** has size $S \times S$ whereas the matrix **B** has size $S \times D$.

The goal of structural controllability, which we employ here, is to use the information contained in **A** to generate a supportable estimate of **B** (and by extension *D*).
This focus allows us to gain insight of the inherent controllability of a network, and the roles of the species that compose it, without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off is that, because of the assumption that ecological communities can be modelled with linear functional responses, structural controllability is not sufficient to fully design a the time-varying control signal *u(t)* that can drive the system from one particular equilibrium to another. 
Nevertheless, the lessons gained when assuming linearity---at both the network and the species level---are a prerequisite for understanding nonlinear control [@Liu2011; @Liu2016].

>>>>

At the core of representing ecological communities as complex systems sits the idea that the state of any given node depends on its state, the state of the nodes it interacts with, the state of the nodes they interact with, and so on. 
These dependencies are given by the structure of the interactions, the dynamic relationships between nodes, and the mechanisms of self-regulation. 
This representation has been very useful for our understanding of ecological communities and might be advantageous for a more formal approach to ecological management based on control theory.

The overall objective of control theory is to be able to steer a system from one state to another in finite time. 
However, in complex systems like ecological communities, the large number of nodes and interactions as well as their non-linear dynamics render its control extremely challenging.
On one theoretical extreme, for instance, any ecological community could be fully controlled if we control the state of every species independently. 
On another, the mechanisms of self-regulation and the multiple feedback cycles found in ecological communities mean it is also theoretically possible to control the whole community by directly modifying the state of just a *single* species [@Cowan2012; @Markovsky2013]. 
Although mathematically correct, neither of these options are practical in real-world applications: the first because it is infeasible to design and implement interventions that modify the abundance of every single species in a community, and the second because the control signal might require unreasonably large amounts of time, energy, or unattainable rapid changes on the state [@Yan2012; @Motter2015].
Here we explore an intermediate point between these two trivial solutions.

In particular, we leverage the principle of species interdependence to find an intermediate set of *driver nodes* to control the network [@Liu2016]. 
When we focus on net interspecific effects, it is possible to identify a minimum number of nodes the control of which can theoretically drive the state of every other node in the network to a desired configuration. 
Conveniently the information necessary to determine this minimum number of driver nodes *D* is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how the species respond to the external input, respectively. 
If *S* is the number of species in the community, the matrix **A** has size $S \times S$ whereas the matrix **B** has size $S \times D$.

The goal of structural controllability, which we employ here, is to use the information contained in **A** to generate a supportable estimate of **B** (and by extension *D*).
This focus allows us to gain insight of the inherent controllability of a network, and the roles of the species that compose it, without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off of our approach is that, because of the assumption of linearity, structural controllability alone does not allow us to fully design the time-varying control signal *u(t)* that can drive the system from one particular equilibrium to another. 
Nevertheless, the lessons gained when assuming linearity---at both the network and the species level---are a prerequisite for eventually understanding nonlinear control [@Liu2011; @Liu2016].
